On mapping errors and seeding heuristics for DNA alignment

### 2019-04-11 MEM.tex
added acknowledgements statement
Incorporated Ruggero's edits to the main text and figures.

This observation is fundamental: in order to estimate the error rate of of
a mapping algorithm, we must know how often the associated seeding
heuristic is off target. For lack of theoretical framework, this
information is missing for nearly % nearly, or all? this is important because if
% there is at least one seeding algorithm that actually estimates correctly the
% probabilities for the seeding heuristics, then all of the rest is much less
% meaningful.
all existing seeding methods, meaning
that current mappers cannot be properly calibrated.

@@

As mentioned above, after the terminator $\Downarrow_i$, we can append
segments with up to $\gamma+i-1$ symbols $\square$. The terminators of
those $\gamma+i$ possible segments are distributed according to the laws
of the arithmetic modulo $n+1$. For $i > 0$ for instance, if the next
segment has no symbol $\square$ then it is terminated by
$\Downarrow_{i-1}$. But the same goes if the segment has $n+1$ symbols
$\square$ (assuming $n+1 < \gamma$). Specifying the transfer matrix thus
involves a substantial amount of bookkeeping. One can check that the
expression of the transfer matrix $M_n(z)$ is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{eqnarray}
H_{i,j}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
\end{eqnarray}
where $\lfloor \ldots \rfloor$ denotes the ``floor'' function.

@@

$M_n(z)$ is too complex to allow computing a closed expression of $M_n(z) \cdot
(I-M_n(z))^{-1}$. We return to the definition $M_n(z) + M_n(z)^2 + \ldots$ and
observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3}, \ldots$ have no influence on
the coefficients of the weighted generating function $a_0, a_1, \ldots, a_k$.

@@

$\tilde{M}_0(z)$ is too complex to allow computing a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is actually easier to
proceed as in section~\ref{sec:example_skip} and to compute the powers of
$\tilde{M}_0(z)$ (using the arithmetic of truncated polynomials) up to a
given threshold value. Since each segment except the tail contains a
mismatch against at least one sequence, we define $\tilde{p}$ as the upper
bound on the probability of a mismatch $\tilde{p} = \max\{b,c,d\}$. The
updated formula (\ref{eq:bound}) now gives an upper bound of the
probability that a read of size $k$ contains $s$ or more mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

@@

We acknowledge the financial support of the Spanish Ministry of Economy and
Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan Nacional
BFU2012-37168), of the CERCA Programme~/~Generalitat de Catalunya, and of the
European Research Council (Synergy Grant 609989).

>>>>

This observation is fundamental: in order to estimate the error rate of of
a mapping algorithm, we must know how often the associated seeding
heuristic is off target. For lack of theoretical framework, this
information is missing, meaning that current mappers cannot be properly
calibrated.

@@

As mentioned above, after the terminator $\Downarrow_i$, we can append
segments with up to $\gamma+i-1$ symbols $\square$. The terminators of
those $\gamma+i$ possible segments are distributed according to the laws
of the arithmetic modulo $n+1$. For $i > 0$ for instance, if the next
segment has no symbol $\square$ then it is terminated by
$\Downarrow_{i-1}$. But the same goes if the segment has $n+1$ symbols
$\square$ (assuming $n+1 < \gamma$). Specifying the transfer matrix thus
involves a substantial amount of bookkeeping. One can check that the
expression of the transfer matrix $M_n(z)$ is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{eqnarray}
H_{i,j}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1},
\end{eqnarray}
where $\lfloor \ldots \rfloor$ denotes the ``floor'' function.

@@

$M_n(z)$ is too complex to allow a closed expression of $M_n(z) \cdot
(I-M_n(z))^{-1}$ to be computed. We return to the definition $M_n(z) +
M_n(z)^2 + \ldots$ and observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3},
\ldots$ have no influence on the coefficients of the weighted generating
function $a_0, a_1, \ldots, a_k$.

@@

$\tilde{M}_0(z)$ is too complex to allow a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$ to be computed. It is
actually easier to proceed as in section~\ref{sec:example_skip} and to
compute the powers of $\tilde{M}_0(z)$ (using the arithmetic of truncated
polynomials) up to a given threshold value. Since each segment except the
tail contains a mismatch against at least one sequence, we define
$\tilde{p}$ as the upper bound on the probability of a mismatch $\tilde{p}
= \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now gives an upper
bound of the probability that a read of size $k$ contains $s$ or more
mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

@@

We acknowledge the financial support of the Spanish Ministry of Economy and
Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan Nacional
BFU2012-37168), of the CERCA Programme~/~Generalitat de Catalunya, and of the
European Research Council (Synergy Grant 609989). R.~C.~ was supported by the
People Programme (Marie Curie Actions) of the European Union's Seventh Framework
Programme (FP7/2007-2013) under REA grant agreement 608959.
